---
title: "Lab 02 - Plastic waste"
author: "Lilly McClendon"
date: "2025-01-23"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2)
```
```{r country-waste-over-3.5}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```
##### Factors related to high plastic waste per capita in Trinidad and Tobago
According to Kanhai and colleagues in 2024, individuals in Trinidad and Tobago use plastics because of their affordable price, need, little belief that they can change their behavior, and convenience. Additionally, Kanhai and colleagues in 2024 found factors such as social norms and convenience are predictors for single-use plastics. Further, even with increasing awareness of recycling, only half of people recycled, potentially due to inaccessibility of recycling. 

#### 1.1 histogram plot of distribution of plastic waster per capita faceted by continent

```{r histogram-plastic-waste-percap-facet-by-country}
ggplot(plastic_waste, aes(x=plastic_waste_per_cap)) +
  geom_histogram() +
  facet_wrap(~continent)
```

#### Continent comparison of plastic waster per capita
Surprisingly it seems like the continents all share similar patterns except for South America which appears to have a drastically smaller plastic waste per capita. 

### Exercise 2

```{r visualizing numerical data density plots }
ggplot(data = plastic_waste,aes(x = plastic_waste_per_cap, shape = continent)) +
  geom_density()
```

```{r desnity curve continents color coded}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, color = continent)) +
  geom_density()
```


```{r density curve continents fill colored}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, color = continent, fill = continent)) +
  geom_density()
```

#### 2.1 Desnity Plots with different Alphas 
```{r density curve continents fill colored transparency}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, color = continent, fill = continent)) +
  geom_density(alpha = .25)
```

#### 2.1 Definition of color and fill by mapping aesthetics and definition of alpha levels as characteristic of plotting geom. 

We defined color and fill as mapping aesthetics because we wanted the color and fill to be based on the values of a variable in the data, in this case continent. We defined alpha levels as characteristics of plotting geom because we wanted to modify how transparent the points were from 0-totally transparent to 1-totally solid. The alpha values were therefore not based on the values of a variable in the data so we defined the alpha level as a characteristic of the plotting geom.


### Exercise 3

```{r visualize plastic data side-by-side box plots}
ggplot(data = plastic_waste,mapping = aes(x = continent, y = plastic_waste_per_cap), title ) +
  geom_boxplot()
```




#### 3.1 Convert side-by-side box plots to violin plots

```{r plastic-waste-violin}
ggplot(plastic_waste, aes(x = continent, y=plastic_waste_per_cap))+
  geom_violin()
```

#### Violin plots compared to box plots 
In addition to reflecting summary statistics like the box plots, a unique feature of violin plots is that you can see the density of each variable (similar to a flipped histogram). Rather than showing bars to represent the data, the distribution itself is plotted turned. So if you look at the violin plot of South America you can see that the data is densest near the 0 plastic waste per capita.        

#### Features of box plots not displayed in violin plots 
In contrast to violin plots, box plots display the specific values of the summary statistics like the quartiles, median, and clearly distinguish outliers. The line inside the box represents the median value, and the bottom horizontal edge of the box represents the first quartile and the top horizontal edge of the box represents the third quartile.     

### Exercise 4

```{r plastic-waste-mismanaged}
# insert code here
```

```{r plastic-waste-mismanaged-continent}
# insert code here
```

```{r plastic-waste-population-total}
# insert code here
```

```{r plastic-waste-population-coastal}
# insert code here
```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r recreate-viz}
# insert code here
```


### References

Kanhai, L. D. K., Keller, E., & Richter, I. (2024). The human dimension of plastic pollution in the Caribbean SIDS of Trinidad and Tobago. Environmental Science & Policy, 159, 103820-. https://doi.org/10.1016/j.envsci.2024.103820


