---
title: "Lab 02 - Plastic waste"
author: "Lilly McClendon"
date: "2025-01-23"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
```

```{r load-data, message=FALSE}
plastic_waste <- read.csv("data/plastic-waste.csv")
```

## Exercises

### Exercise 1

```{r plastic-waste-continent}
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_histogram(binwidth = 0.2) + 
  labs(title = "Plastic waste per capita in 2010")
```
```{r country-waste-over-3.5}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.5)
```
##### Factors related to high plastic waste per capita in Trinidad and Tobago
According to Kanhai and colleagues in 2024, individuals in Trinidad and Tobago use plastics because of their affordable price, need, little belief that they can change their behavior, and convenience. Additionally, Kanhai and colleagues in 2024 found factors such as social norms and convenience are predictors for single-use plastics. Further, even with increasing awareness of recycling, only half of people recycled, potentially due to inaccessibility of recycling. 

#### 1.1 histogram plot of distribution of plastic waster per capita faceted by continent

```{r histogram-plastic-waste-percap-facet-by-country}
ggplot(plastic_waste, aes(x=plastic_waste_per_cap)) +
  geom_histogram() +
  facet_wrap(~continent) +
  labs(title = "Plastic waste per capita by country")
```

##### Continent comparison of plastic waster per capita
Surprisingly it seems like the continents all share similar patterns except for South America which appears to have a drastically smaller plastic waste per capita. 

### Exercise 2

#### Density plots 
```{r visualizing numerical data density plots }
ggplot(data = plastic_waste, aes(x = plastic_waste_per_cap)) +
  geom_density() + 
  labs(title = "Plastic waste per capita by continent")
```


```{r visualizing numerical data density plots all continents }
ggplot(data = plastic_waste,aes(x = plastic_waste_per_cap, shape = continent)) +
  geom_density() + 
  labs(title = "Plastic waste per capita by continent")
```

#### Density plots color coded by continent

```{r desnity curve continents color coded}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, color = continent)) +
  geom_density() + 
  labs(title = "Plastic waste per capita by continent")
```

#### Density plots filled curves
```{r density curve continents fill colored}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, color = continent, fill = continent)) +
  geom_density() + 
  labs(title = "Plastic waste per capita by continent")
```

#### 2.1 Desnity Plots with different Alphas 
```{r density curve continents fill colored transparency}
ggplot(data = plastic_waste, mapping = aes(x = plastic_waste_per_cap, color = continent, fill = continent)) +
  geom_density(alpha = .25) + 
  labs(title = "Plastic waste per capita by continent")
```

##### 2.1 Definition of color and fill by mapping aesthetics and definition of alpha levels as characteristic of plotting geom. 

We defined color and fill as mapping aesthetics because we wanted the color and fill to be based on the values of a variable in the data, in this case continent. We defined alpha levels as characteristics of plotting geom because we wanted to modify how transparent the points were from 0-totally transparent to 1-totally solid. The alpha values were therefore not based on the values of a variable in the data so we defined the alpha level as a characteristic of the plotting geom.


### Exercise 3

```{r visualize plastic data side-by-side box plots}
ggplot(data = plastic_waste,mapping = aes(x = continent, y = plastic_waste_per_cap)) +
  geom_boxplot() + 
  labs(title = "Plastic waste per capita vs. continent")
```



#### 3.1 Convert side-by-side box plots to violin plots

```{r plastic-waste-violin}
ggplot(plastic_waste, aes(x = continent, y=plastic_waste_per_cap))+
  geom_violin() + 
  labs(title = "Plastic waste per capita vs. continent")
```

##### Violin plots compared to box plots 
In addition to reflecting summary statistics like the box plots, a unique feature of violin plots is that you can see the density of each variable (similar to a flipped histogram). Rather than showing bars to represent the data, the distribution itself is plotted turned. So if you look at the violin plot of South America you can see that the data is densest near the 0 plastic waste per capita.        

##### Features of box plots not displayed in violin plots 
In contrast to violin plots, box plots display the specific values of the summary statistics like the quartiles, median, and clearly distinguish outliers. The line inside the box represents the median value, and the bottom horizontal edge of the box represents the first quartile and the top horizontal edge of the box represents the third quartile.     

### Exercise 4

#### 4.1 Visualize relationship between plastic waste per capita and mismanaged plastic waste per capita with a scatterplot
```{r plastic-waste-mismanaged-scatterplot}
ggplot(plastic_waste, aes(x = mismanaged_plastic_waste_per_cap, y = plastic_waste_per_cap)) + 
  geom_point() + 
  labs(title = "Plastic waste per capita vs. mismanaged plastic waste per capita")
```

##### 4.1 Description of relationship (scatterplot of relationship between plastic waste per capita and mismanaged plastic waste per capita)

This scatterplot shows a weak, positive linear relationship between plastic waste per capita and mismanaged plastic waste per capita and one clear outlier in the data. 

#### 4.2 Color scatterplot points by continent 

```{r plastic-waste-mismanaged-continent}
ggplot(plastic_waste, aes(x = mismanaged_plastic_waste_per_cap, y = plastic_waste_per_cap, color = continent)) + 
  geom_point() + 
  labs(title = "Plastic waste per capita vs. mismanaged plastic waste per capita")
```

##### 4.2 Distinctions between continents with how plastic waste per capita and mismanaged plastic waste per capita are associated

It seems that there is a stronger linear relationship for Africa compared to other countries. For example it seems as though North America has weaker association.  

#### 4.3 Relationship between plastic waste per capita and total population

```{r plastic-waste-population-total}
ggplot(plastic_waste, aes(x = total_pop, y = plastic_waste_per_cap, color = continent)) + 
  geom_point() + 
  labs(title = "Plastic waste per capita vs. total population")
```

#### 4.3 Relationship between plastic waste per capita and coastal population

```{r plastic-waste-population-coastal}
ggplot(plastic_waste, aes(x = coastal_pop , y = plastic_waste_per_cap, color = continent)) + 
  geom_point()
  labs(title = "Plastic waste per capita vs. coastal population")
```

##### 4.3 
The relationship between plastic waste per capita and total population and the relationship between plastic waste per capita and coastal population appear to be similiarly linearly associated. I would say that one is not more strongly linearly associated than the other. 

### Exercise 5


```{r recreate-viz}
plastic_waste %>%
  filter(plastic_waste_per_cap > 3.0)

ggplot(plastic_waste, aes(x = (coastal_pop/total_pop), y =plastic_waste_per_cap)) + 
  geom_point(aes(color = continent)) +
  xlim(0, 1.75) + 
  ylim(0, 0.7) + 
  labs(title = "Plastic waste vs. coastal population proportion", subtitle = "by continent", x = "Coastal population proportion (Coastal/total population)", y = "Plastic waste per capita") +
  scale_color_viridis_d() + 
  geom_smooth(method = "loess", col = "black")
```

##### 5.1 Overall there is a weak positive association between plastic waste per capita and coastal population proportion. It appears that as there is a higher density of popualtion at the coast, there is in increase in plastic waste per capita. It may be that people who reside on the coast are more likely to consume more single-use plastic.  

### References

Kanhai, L. D. K., Keller, E., & Richter, I. (2024). The human dimension of plastic pollution in the Caribbean SIDS of Trinidad and Tobago. Environmental Science & Policy, 159, 103820-. https://doi.org/10.1016/j.envsci.2024.103820


